Theory about Normal Basis on $F(2^k)$

### 2016-12-11 dissertation/prelim_GF.tex
dissertation v2:milestone
dissertation v3:milestone

\chapter{Galois Field and Sequential Arithmetic Circuits}
\label{ch:prelim_GF}
This chapter provides a mathematical background for understanding 
finite fields (Galois fields) and explains how to design Galois field arithmetic circuits.
We first introduce the mathematical concepts of groups, rings, fields, and 
polynomials. 
We then apply these concepts to create Galois field arithmetic functions and 
explain how to map them to a Boolean circuit implementation.
Additonally, we introduce a special type of sequential arithmetic hardware based on normal basis, as well
as the normal basis theory behind the designing such hardware.

The material is referred from \cite{galois_field:mceliece, ftheory:2006, ff:1997} for Galois field concepts, 
\cite{mastro:1989, PT:1985, acar:1998, wu:2002, Knezevic:2008} for hardware design over Galois fields 
and previous work by {\it Lv} \cite{lv:phd}.
Normal basis theory in this section refers to \cite{normal_book, gao:phd_normal_basis} and sequential
normal basis arithmetic hardware designs come from \cite{mullinONB,MasseyOmura,agnew1991implementation, RHmulti}.

\section{Commutative Algebra}
\label{sec:algebra}
\subsection{Group, ring and field}
\begin{Definition}
An {\bf abelian group} is a set $\mathbb{S}$ with a binary operation $'+'$
which satisfies the following properties: 
\begin{itemize}
\item {\it Closure Law:} For every $a, b \in \mathbb{S}, a + b \in \mathbb{S}$  
\item {\it Associative Law:} For every $a, b, c \in \mathbb{S}, (a + b) + c = a + (b + c)$
\item {\it Commutativity:} For every $a, b \in \mathbb{S}, a + b = b + a$. 
\item {\it Additive Identity:} There is an identity element $0 \in \mathbb{S}$
such that for all $a \in \mathbb{S};$ $a + 0 = a$.
\item {\it Additive Inverse:} If $a \in \mathbb{S}$, then there is an
element $a^{-1} \in \mathbb{S}$ such that $ a + a^{-1} = 0$.
\end{itemize}
\end{Definition}

The set of integers $\mathbb{Z}$ forms an abelian group under the addition operation.

\begin{Definition}
Given a set $\mathbb{R}$ with two binary operations, $'+'$ and $'\cdot'$, 
and element $0 \in \mathbb{R}$, the system $\mathbb{R}$ is called a {\bf commutative ring with unity} if the following properties hold:
\begin{itemize}
\item $\mathbb{R}$ forms an abelian group under the '+' operation with additive identity element $0$.
\item {\it Multiplicative Distributive Law}: For all $a, b, c \in$ $\mathbb{R}$, $a\cdot (b + c) = a\cdot b + a\cdot c$.
\item {\it Multiplicative Associative Law}: For every $a, b, c\in \mathbb{R}$, $a\cdot (b\cdot c) = (a\cdot b)\cdot c$. 
\item {\it Multiplicative Commutative Law}: For every $a,b \in \mathbb{R}$, $a\cdot b = b\cdot a$
\item {\it Identity Element}: There exists an element $1 \in$ $\mathbb{R}$ 
such that for all $a \in \mathbb{R}$, $a\cdot 1 = a =1\cdot a$
\end{itemize}
\end{Definition}

@@

\begin{Definition}
A {\bf field} $\mathbb{F}$ is a commutative ring with unity, where every
non-zero element in $\mathbb{F}$ has a multiplicative inverse; i.e. $\forall
a \in \mathbb{F} - \{0\}$, $\exists \hat{a} \in \mathbb{F}$ such that $ a \cdot
\hat{a} = 1$.
\end{Definition}

@@

\begin{Definition} 
A {\bf Galois field}, denote $\Fq$, is a field with a finite
number of elements, $q$. The number of elements $q$ of the Galois field is
a power of a prime integer, i.e. $q = p^k$, where $p$ is a prime
integer, and $k \geq 1$. Thus a Galois field can also be denoted as 
$\F_{p^{k}}$.
\end{Definition}

@@

Algebraic extensions of the binary field $\F_{2}$  
are generally termed as {\it binary extension fields} $\Fkk$.
Where elements in $\F_2$ can only represent $1$ bit, elements in $\Fkk$ 
represent a $k$-bit vector.
This allows them to be widely used in digital hardware applications.
In order to construct a Galois field of the form $\Fkk$, 
an {\bf irreducible polynomial} is required:
\begin{Definition}
A polynomial $P(x) \in \mathbb{F}_{2}\left[x\right]$, i.e. the set of all 
polynomials in $x$ with coefficients in $\F_2$, is {\bf irreducible} 
if $P(x)$ is non-constant with degree $k$ and cannot be 
factored into a product of polynomials of lower degree in $\mathbb{F}_2[x]$.
\end{Definition}

Therefore, the polynomial $P(x)$ with degree $k$ is irreducible over 
$\mathbb{F}_{2}$ if and only if it has no roots in $\mathbb{F}_{2}$,
i.e if $\forall a \in \mathbb{F}_{2}$, $P(a)\neq 0$.
For example, $x^2+x+1$ is an irreducible polynomial over $\mathbb{F}_{2}$
because it has no solutions in $\mathbb{F}_{2}$, i.e. $(0)^2+(0)+1=1\neq0$ 
and $(1)^2+(1)+1=1\neq0$ over $\F_2$.
Irreducible polynomials exist for any degree $\geq 2$ in $\mathbb{F}_2[x]$.

Given an irreducible polynomial $P(x)$ of degree $k$ in the polynomial ring 
$\mathbb{F}_2[x]$, we can construct a binary extension field 
$\mathbb{F}_{2^k} \equiv \mathbb{F}_2[x] \pmod{P(x)}$.
Let $\alpha$ be a root of $P(x)$, i.e., $P(\alpha)=0$.
Since $P(x)$ is irreducible over
$\mathbb{F}_2[x]$, $\alpha \notin \mathbb{F}_2$. 
Instead, $\alpha$ is an element in $\mathbb{F}_{2^k}$. 
Any element $A \in \mathbb{F}_{2^k}$ is then represented as: 
\begin{equation}\label{rep:poly}
A= \sum_{i=0}^{k-1} (a_i \cdot \alpha^i) = a_0 + a_1\cdot\alpha + \cdots + a_{k-1}\cdot \alpha^{k-1}\nonumber
\end{equation}
where $a_i \in \mathbb{F}_2$ are the coefficients and $P(\alpha)=0$.

@@

\begin{Example}\label{exp:1}
Let us construct $\mathbb{F}_{2^4}$ as $\mathbb{F}_2[x] \pmod{ P(x)}$, where
$P(x)=x^4+x^3+1 \in \mathbb{F}_2[x]$ is an irreducible polynomial of degree $k=4$. 
Let $\alpha$ be the root of $P(x)$, i.e. $P(\alpha)=0$.

@@

\begin{Theorem}
$\F_{2^n}\subset\F_{2^m}$ iff $n \mid m$, i.e. if $n$ divides $m$.
\end{Theorem}

@@

\section{Normal Basis Multiplier over Galois Field}
From algebraic view, a field is a space, and field elements are points in the space. Those elements can be 
represented with unique coordinates, which requires the pre-definition of a basis vector. In this
section, we discuss a special basis called normal basis, as well as the advantages adopting it in 
GF operations esp. multiplication.
\subsection{Normal Basis}
Given a Galois field (GF) $\Fkk$ is a finite field with  $2^k$ elements and characteristic equals to 2.
Its elements can be written in polynomials of $\alpha$, when there is an irreducible polynomial $p(\alpha)$
defined.

If we use a basis $\{1,\alpha,\alpha^2,\alpha^3,\dots,\alpha^{k-1}\}$, we can easily transform polynomial representations
to binary bit-vector representations by recording the coefficients. For example, for elements in
$\F_{2^4}$, the results are shown in table \ref{tab:gfelement}, column ``Polynomial".

@@

Basis $\{1,\alpha,\alpha^2,\alpha^3,\dots,\alpha^{k-1}\}$ is called {\bf standard basis} (StdB), which results in
a straightforward representation for elements, and operations of elements such as addition and subtraction.
The addition/subtraction of GF elements in StdB follows the rules of polynomial addition/subtraction
where coefficients belong to $\mathbb F_2$. In other words, using the definition of {\it exclusive or} (XOR) in
Boolean algebra, element $A$ add/subtract by element $B$ in StdB is defined as
\begin{align}\label{eqn:StdB}
A+B = A-B &= (a_0,a_1,\dots,a_{k-1})_{StdB} \xor (b_0,b_1,\dots,b_{k-1})_{StdB} \nonumber\\
&=(a_0\oplus b_0, a_1\oplus b_1,\dots,a_{k-1}\oplus b_{k-1})_{StdB} 
\end{align}

\subsection{Multiplication using Normal Basis}
Besides addition/subtraction, multiplication is also very common in arithmetic circuit design.
The multiplication of GF elements in $\Fkk$ in StdB follows the rule of polynomial multiplication.
However, it will result in $O(k^2)$ bitwise operations. In other words, if we implement GF multiplication
in bit-level logic circuit, it will contain $O(k^2)$ gates. When the datapath size $k$ is large,
the area and delay of circuit will be costly.

In order to lower down the complexity of arithmetic circuit design, Massey and Omura \cite{MasseyOmura} % ref 7 in RH paper
use a new basis to represent GF elements, which is called {\bf normal basis} (NB).
A normal basis over $\Fkk$ is written in the form of
\begin{equation*}
N.B. ~~~ \N = \{\beta,\beta^2,\beta^4,\beta^8,\dots,\beta^{2^{k-1}}\}
\end{equation*}

{\bf Normal element} $\beta$ is an element from the field which is used to construct the normal basis,
and can be represent as a power of primitive element $\alpha$: 
\begin{equation*}
\beta = \alpha^t, ~~~ 1\leq t<2^k
\end{equation*}
Exponent $t$ takes value in the given range when $\N$ fulfills the definition of a basis.

Respectively, a field element in NB representation is actually
\begin{align*}
A &= (a_0,a_1,\dots,a_{k-1})_{NB} \\
  &= a_0\beta+a_1\beta^2+\cdots+a_{k-1}\beta^{2^{k-1}} \\
  &= \sum_{i=0}^{k-1} a_i\beta^{2^i}
\end{align*}

According to the definition, a normal basis is a vector where the next entry is the square of the former one.
We note that the vector is cyclic, i.e. $\beta^{2^k} = \beta$ due to {\it Fermat's little theorem}.

The addition and subtraction of elements in NB representation are similar to equation \ref{eqn:StdB}.
However, what makes NB powerful is its ease of implementation when doing multiplications and exponentiations.
The following lemmas and examples illustrate this fabulous property very well.
\begin{Lemma}[Square of NB]
\label{lem:squareNB}
In $\Fkk$, 
\begin{equation*}
(a+b)^2 = a^2 + b^2
\end{equation*}
According to the \textbf{binomial theorem}, it can be extended to
\begin{align*}
\beta^2 =&(b_0\beta + b_1\beta^2 + b_2\beta^4 + \dots + b_{k-1}\beta^{2^{k-1}})^2 \\
=& b_0^2\beta^2 + b_1^2\beta^4 + b_2^2\beta^8 + \dots + b_{k-1}^2\beta \\
=& b_{k-1}^2\beta + b_0\beta^2 + b_1\beta^4 + \dots + b_{k-2}\beta^{2^{k-1}}
\end{align*}
\end{Lemma}
This lemma concludes that the square of an element in NB equals to a simple right-cyclic shift of the bit-vector.
Obviously, StdB representation does not have this benefit.

@@

This example shows that convenience to use NB when computing $2^k$ power of an element.
Multiplication is a bit complicated than squaring; but when it is decomposed as bit-wise
operations, the property in lemma \ref{lem:squareNB} can be well utilized.

@@

Assume the most significant bit (MSB) of the product can be represented by a function $f_{mult}$: 
\begin{equation}
\label{eqn:multiMSB}
c_{k-1} = f_{mult}(a_0, a_1, \dots, a_{k-1}; b_0, b_1, \dots, b_{k-1})
\end{equation}
Before discussing the details of the function $f_{mult}$, we can take 
a square on both side of equation \ref{eqn:multiMSB}, i.e. $C^2 = A^2\times B^2$.
Obviously, using the property in lemma \ref{lem:squareNB}, the original second most significant bit 
becomes the new MSB because of right-cyclic shifting. 
Concretely, 
$$(c_{k-1},c_0,c_1,\dots,c_{k-2}) = (a_{k-1},a_0,a_1,\dots,a_{k-2})\times(b_{k-1},b_0,b_1,\dots,b_{k-2})$$
Note $A^2, B^2$ and $C^2$ still belong to $\Fkk$, thus as a universal function implementing MSB multiplication
over $\Fkk$, $f_{mult}$ still remains the same. As a result, the new MSB can be written as 
\begin{equation}
\label{eqn:shiftMSB}
c_{k-2} = f_{mult}(a_{k-1}, a_0, a_1, 
\dots, a_{k-2}; b_{k-1}, b_0, b_1, \dots, b_{k-2})
\end{equation}
Similarly, if we take a square again on the new equation, we can get $c_{k-3}$.
Successively we can derive all bits of product $C$ using the same function $f_{mult}$, and the only
adjustment we need to make is to right-cyclic shift 2 operands by 1 bit each time.
\end{Example}

@@

If we want to make the complexity of $f_{mult}$ lower than $O(k^2)$, then the best choice is to try out 
linear functions. As we know, matrix multiplication can simulate all possible combinations of 
linear functions. 
% (which is also the reason it is used as basic model to simulate the behavior of a neuron
% in neural network machine learning algorithms).
Imagine $A$ is a $k$-bit row vector and $B$ is a $k$-bit 
column vector, then the single bit product can be written as the product of matrix multiplication
$$c_{l} = A\times M\times B$$
where 
\begin{align*}
A &= (a_0,a_1,\dots,a_{k-1})\\
B &= \begin{pmatrix}
b_0 \\
b_1 \\
\vdots \\
b_{k-1}
\end{pmatrix}
\end{align*}
Moreover, $M$ is a $k\times k$ square matrix. If we can find $M$, we obtain the design of the multiplier.

@@

When taking different bits $l$ of the product in equation \ref{eqn:def_lambda}, 
we obtain a series of conjugate matrices of $M$. 
Which means instead of shifting operands $A$ and $B$, we can also shift the 
matrix.

More specifically, we denote the matrix by \emph{$l$-th $\lambda$-Matrix} as 
$$c_l = A \times M^{(l)} \cdot B^T$$
Meanwhile, the operator shifting rule in equation \ref{eqn:shiftMSB} still holds. Then we have relation 
$$c_{l-1} = A \cdot M^{(l-1)} \cdot B^T = shift(A) \cdot M^{(l)} \cdot shift(B)^T$$ which means
by right and down cyclically shifting $M^{(l-1)}$, we can get $M^{(l)}$.

@@

If we generalize the definition and explore the nature of $\lambda$-Matrix, it is defined as cross-product terms from multiplication, which is 
\begin{equation}
Product~vector~C = (\sum_{i=0}^{k-1}a_i\beta^{2^i})(\sum_{j=0}^{k-1}b_j\beta^{2^j}) = \sum_{i=0}^{k-1}\sum_{j=0}^{k-1}a_ib_j\beta^{2^i}\beta^{2^j}
\end{equation}
The expressions $\beta^{2^i}\beta^{2^j}$ are referred to as cross-product terms, and can be represented by
NB, i.e.
\begin{equation}
\beta^{2^i}\beta^{2^j} = \sum_{l=0}^{k-1}\lambda_{ij}^{(l)}\beta^{2^l}, \ \ \lambda_{ij}^{(l)} \in \mathbb F_2.
\end{equation}
Substitution yields, result is an expression for l-th digit of product as showed in equation \ref{eqn:multiMSB}:
\begin{equation}
c_l = \sum_{i=0}^{k-1}\sum_{j=0}^{k-1}\lambda_{ij}^{(l)}a_ib_j
\end{equation}
$\lambda_{ij}^{(l)}$ is the entry with coordinate $(i,j)$ in $l$-th $\lambda$-Matrix.

The $\lambda$-Matrix can be implemented with XOR and AND gates in circuit design.
The very naive implementation requires $O(C_N)$ gates, where $C_N$ is the number of
nonzero entries in $\lambda$-Matrix.
There usually exists multiple NBs in $\Fkk, k>3$. If we employ a random NB, there is no mathematical
guarantee that $C_N$ has bound $o(k)$. However,
Mullin et al. \cite{mullinONB} % This citation is valid
proves that in certain GF $\mathbb F_{p^{k_{opt}}}$, there always exists at least one NB such that 
its corresponding $\lambda$-Matrix has $C_N = 2n-1$ nonzero entries. A basis with this property is
called optimal normal basis (ONB), details are introduced in appendix \ref{append:ONB}.

@@

% After fixing this, add a whole-piece StdB vs NB example, list their cost as the conclusion
\subsection{Comparison between Standard Basis and Normal Basis}
At the end of this section, a detailed example is used to make a comparison between StdB multiplication
and NB multiplication.
\begin{Example}[Rijndael's finite field]
Rijndael uses a characteristic 2 finite field with 256 elements, which can also be called the GF $\mathbb F_{2^8}$.
Let us define the primitive element $\alpha$ using irreducible polynomial 
$\alpha^8+\alpha^7+\alpha^6+\alpha^4+\alpha^2+\alpha+1$. Coincidently, $\alpha$ is also a normal element,
i.e. $\beta = \alpha$ can construct a NB $\{\alpha,\alpha^2,\alpha^4,\alpha^8,\alpha^{16},\alpha^{32},\alpha^{64},
\alpha^{128}\}$.

@@

The final remainder is $1$, i.e. the product equals to 1 in StdB.

>>>>

\chapter{Finite Fields and Sequential Arithmetic Circuits}
\label{ch:prelim_GF}
This chapter provides a mathematical background for understanding 
finite fields (Galois fields) and explains how to design Galois field (GF) arithmetic circuits.
We first introduce the mathematical concepts of groups, rings, fields, and 
polynomials. 
We then apply these concepts to create Galois field arithmetic functions and 
explain how to map them to a Boolean circuit implementation.
Additionally, we introduce a special type of sequential arithmetic hardware based on normal basis, as well
as the normal basis theory behind the designing such hardware.

The material is referred from \cite{galois_field:mceliece, ftheory:2006, ff:1997} for Galois field concepts, 
\cite{mastro:1989, PT:1985, acar:1998, wu:2002, Knezevic:2008} for hardware design over Galois fields 
and previous work by Lv \cite{lv:phd}.
Normal basis theory in this section is referred from \cite{normal_book, gao:phd_normal_basis} and sequential
normal basis arithmetic hardware designs come from \cite{mullinONB,MasseyOmura,agnew1991implementation, RHmulti}.

\section{Commutative Algebra}
\label{sec:algebra}
\subsection{Group, Ring and Field}
\begin{Definition}
An {\bf Abelian group} is a set $\mathbb{S}$ with a binary operation $'+'$
which satisfies the following properties: 
\begin{itemize}
\item {\it Closure Law:} For every $a, b \in \mathbb{S}, a + b \in \mathbb{S}$  
\item {\it Associative Law:} For every $a, b, c \in \mathbb{S}, (a + b) + c = a + (b + c)$
\item {\it Commutativity:} For every $a, b \in \mathbb{S}, a + b = b + a$. 
\item {\it Additive Identity:} There is an identity element $0 \in \mathbb{S}$
such that for all $a \in \mathbb{S};$ $a + 0 = a$.
\item {\it Additive Inverse:} If $a \in \mathbb{S}$, then there is an
element $a^{-1} \in \mathbb{S}$ such that $ a + a^{-1} = 0$.
\end{itemize}
\end{Definition}

The set of integers $\mathbb{Z}$ forms an Abelian group under the addition operation.

\begin{Definition}
Given a set $\mathbb{R}$ with two binary operations, $'+'$ and $'\cdot'$, 
and element $0 \in \mathbb{R}$, the system $\mathbb{R}$ is called a {\bf commutative ring with unity} if the following properties hold:
\begin{itemize}
\item $\mathbb{R}$ forms an Abelian group under the '+' operation with additive identity element $0$.
\item {\it Multiplicative Distributive Law}: For all $a, b, c \in$ $\mathbb{R}$, $a\cdot (b + c) = a\cdot b + a\cdot c$.
\item {\it Multiplicative Associative Law}: For every $a, b, c\in \mathbb{R}$, $a\cdot (b\cdot c) = (a\cdot b)\cdot c$. 
\item {\it Multiplicative Commutative Law}: For every $a,b \in \mathbb{R}$, $a\cdot b = b\cdot a$
\item {\it Identity Element}: There exists an element $1 \in$ $\mathbb{R}$ 
such that for all $a \in \mathbb{R}$, $a\cdot 1 = a =1\cdot a$
\end{itemize}
\end{Definition}

@@

\begin{Definition}
A {\bf field} $\mathbb{F}$ is a commutative ring with unity, where every
non-zero element in $\mathbb{F}$ has a multiplicative inverse; {\it i.e.} $\forall
a \in \mathbb{F} - \{0\}$, $\exists \hat{a} \in \mathbb{F}$ such that $ a \cdot
\hat{a} = 1$.
\end{Definition}

@@

\begin{Definition} 
A {\bf Galois field}, denote $\Fq$, is a field with a finite
number of elements, $q$. The number of elements $q$ of the field is
a power of a prime integer, {\it i.e.} $q = p^k$, where $p$ is a prime
integer, and $k \geq 1$. Thus a Galois field can also be denoted as 
$\F_{p^{k}}$.
\end{Definition}

@@

Algebraic extensions of the binary field $\F_{2}$  
are generally termed as {\it binary extension fields} $\Fkk$.
Where elements in $\F_2$ can only represent $1$ bit, elements in $\Fkk$ 
represent a $k$-bit vector.
This allows them to be widely used in digital hardware applications.
In order to construct a Galois field of the form $\Fkk$, 
an {\bf irreducible polynomial} is required:
\begin{Definition}
A polynomial $P(x) \in \mathbb{F}_{2}\left[x\right]$, {\it i.e.} the set of all 
polynomials in $x$ with coefficients in $\F_2$, is {\bf irreducible} 
if $P(x)$ is non-constant with degree $k$ and cannot be 
factored into a product of polynomials of lower degree in $\mathbb{F}_2[x]$.
\end{Definition}

Therefore, the polynomial $P(x)$ with degree $k$ is irreducible over 
$\mathbb{F}_{2}$ if and only if it has no roots in $\mathbb{F}_{2}$,
i.e if $\forall a \in \mathbb{F}_{2}$, $P(a)\neq 0$.
For example, $x^2+x+1$ is an irreducible polynomial over $\mathbb{F}_{2}$
because it has no solutions in $\mathbb{F}_{2}$, {\it i.e.} $(0)^2+(0)+1=1\neq0$ 
and $(1)^2+(1)+1=1\neq0$ over $\F_2$.
Irreducible polynomials exist for any degree $\geq 2$ in $\mathbb{F}_2[x]$.

Given an irreducible polynomial $P(x)$ of degree $k$ in the polynomial ring 
$\mathbb{F}_2[x]$, we can construct a binary extension field 
$\mathbb{F}_{2^k} \equiv \mathbb{F}_2[x] \pmod{P(x)}$.
Let $\alpha$ be a root of $P(x)$, {\it i.e.} $P(\alpha)=0$.
Since $P(x)$ is irreducible over
$\mathbb{F}_2[x]$, $\alpha \notin \mathbb{F}_2$. 
Instead, $\alpha$ is an element in $\mathbb{F}_{2^k}$. 
Any element $A \in \mathbb{F}_{2^k}$ is then represented as: 
\begin{equation}\label{rep:poly}
A= \sum_{i=0}^{k-1} (a_i \cdot \alpha^i) = a_0 + a_1\cdot\alpha + \cdots + a_{k-1}\cdot \alpha^{k-1}\nonumber
\end{equation}
where $a_i \in \mathbb{F}_2$ are the coefficients and $P(\alpha)=0$.

@@

\begin{Example}\label{exp:1}
Let us construct $\mathbb{F}_{2^4}$ as $\mathbb{F}_2[x] \pmod{ P(x)}$, where
$P(x)=x^4+x^3+1 \in \mathbb{F}_2[x]$ is an irreducible polynomial of degree $k=4$. 
Let $\alpha$ be the root of $P(x)$, {\it i.e.} $P(\alpha)=0$.

@@

\begin{Theorem}
$\F_{2^n}\subset\F_{2^m}$ iff $n \mid m$, {\it i.e.} if $n$ divides $m$.
\end{Theorem}

@@

\section{Normal Basis Multiplier over Galois Field}
From an algebraic perspective, a field is a space, and field elements are points in the space. Those elements can be 
represented with unique coordinates, which requires the pre-definition of a basis vector. In this
section, we discuss a special basis called normal basis, as well as the advantages adopting it in 
GF operations, especially multiplication.
\subsection{Normal Basis}
Given a Galois field $\Fkk$ is a finite field with  $2^k$ elements and characteristic equals to 2.
Its elements can be written in polynomials of $\alpha$, when there is an irreducible polynomial $p(\alpha)$
defined.

If we use a basis $\{1,\alpha,\alpha^2,\alpha^3,\dots,\alpha^{k-1}\}$, we can easily transform polynomial representations
to binary bit-vector representations by recording the coefficients. For example, for elements in
$\F_{2^4}$, the results are shown in Table \ref{tab:gfelement}, column ``Polynomial".

@@

The basis $\{1,\alpha,\alpha^2,\alpha^3,\dots,\alpha^{k-1}\}$ is called a {\bf standard basis} (StdB), which results in
a straightforward representation for elements, and operations of elements such as addition and subtraction.
The addition/subtraction of GF elements in StdB follows the rules of polynomial addition/subtraction
where coefficients belong to $\mathbb F_2$. In other words, using the definition of {\it exclusive or} (XOR) in
Boolean algebra, element $A$ add/subtract by element $B$ in StdB is defined as
\begin{align}\label{eqn:StdB}
A+B = A-B &= (a_0,a_1,\dots,a_{k-1})_{StdB} \xor (b_0,b_1,\dots,b_{k-1})_{StdB} \nonumber\\
&=(a_0\oplus b_0, a_1\oplus b_1,\dots,a_{k-1}\oplus b_{k-1})_{StdB} 
\end{align}

\subsection{Multiplication using Normal Basis}
Besides addition/subtraction, multiplication is also very common in arithmetic circuit design.
The multiplication of GF elements in $\Fkk$ in StdB follows the rule of polynomial multiplication.
However, it will result in $O(k^2)$ bitwise operations. In other words, if we implement GF multiplication
in a bit-level logic circuit, it will contain $O(k^2)$ gates. When the datapath size $k$ is large,
the area and delay of circuit will be costly.

In order to lower the complexity of arithmetic circuit design, Massey and Omura \cite{MasseyOmura} % ref 7 in RH paper
use a new basis to represent GF elements, which is called a {\bf normal basis} (NB).
A normal basis over $\Fkk$ is written in the form of
\begin{equation*}
N.B. ~~~ \N = \{\beta,\beta^2,\beta^4,\beta^8,\dots,\beta^{2^{k-1}}\}
\end{equation*}

{\bf Normal element} $\beta$ is an element from the field which is used to construct the normal basis,
and can be represented as a power of the primitive element $\alpha$: 
\begin{equation*}
\beta = \alpha^t, ~~~ 1\leq t<2^k
\end{equation*}
Exponent $t$ takes values in the given range when $\N$ fulfills the definition of a basis.

Correspondingly, a field element in NB representation is actually
\begin{align*}
A &= (a_0,a_1,\dots,a_{k-1})_{NB} \\
  &= a_0\beta+a_1\beta^2+\cdots+a_{k-1}\beta^{2^{k-1}} \\
  &= \sum_{i=0}^{k-1} a_i\beta^{2^i}
\end{align*}

According to the definition, a normal basis is a vector where the next entry is the square of the former one.
We note that the vector is cyclic, {\it i.e.} $\beta^{2^k} = \beta$ due to {\it Fermat's little theorem}.

The addition and subtraction of elements in NB representation are similar to Equation \ref{eqn:StdB}.
However, what makes NB powerful is its ease of implementation when doing multiplications and exponentiations.
The following lemmas and examples illustrate this fabulous property very well.
\begin{Lemma}[Square of NB]
\label{lem:squareNB}
In $\Fkk$, 
\begin{equation*}
(a+b)^2 = a^2 + b^2
\end{equation*}
According to the \textbf{binomial theorem}, it can be extended to
\begin{align*}
\beta^2 =&(b_0\beta + b_1\beta^2 + b_2\beta^4 + \dots + b_{k-1}\beta^{2^{k-1}})^2 \\
=& b_0^2\beta^2 + b_1^2\beta^4 + b_2^2\beta^8 + \dots + b_{k-1}^2\beta \\
=& b_{k-1}^2\beta + b_0\beta^2 + b_1\beta^4 + \dots + b_{k-2}\beta^{2^{k-1}}
\end{align*}
\end{Lemma}
This lemma concludes that the square of an element in NB equals to a simple {\it right-cyclic shift} of the bit-vector.
Obviously, StdB representation does not have this benefit.

@@

This example shows that convenience to use NB when computing $2^k$ power of an element.
Multiplication is more complicated than squaring; but when it is decomposed as bit-wise
operations, the property in Lemma \ref{lem:squareNB} can be well utilized.

@@

Assume the most significant bit (MSB) of the product can be represented by a function $f_{mult}$: 
\begin{equation}
\label{eqn:multiMSB}
c_{k-1} = f_{mult}(a_0, a_1, \dots, a_{k-1}; b_0, b_1, \dots, b_{k-1})
\end{equation}
Before discussing the details of the function $f_{mult}$, we can 
square both sides of Equation \ref{eqn:multiMSB}, {\it i.e.} $C^2 = A^2\times B^2$.
Obviously, using the property in Lemma \ref{lem:squareNB}, the original second most significant bit 
becomes the new MSB because of right-cyclic shifting. 
Concretely, 
$$(c_{k-1},c_0,c_1,\dots,c_{k-2}) = (a_{k-1},a_0,a_1,\dots,a_{k-2})\times(b_{k-1},b_0,b_1,\dots,b_{k-2})$$
Note $A^2, B^2$ and $C^2$ still belong to $\Fkk$, thus as a universal function implementing MSB multiplication
over $\Fkk$, $f_{mult}$ still remains the same. As a result, the new MSB can be written as 
\begin{equation}
\label{eqn:shiftMSB}
c_{k-2} = f_{mult}(a_{k-1}, a_0, a_1, 
\dots, a_{k-2}; b_{k-1}, b_0, b_1, \dots, b_{k-2})
\end{equation}
Similarly, if we take a square again on the new equation, we can get $c_{k-3}$.
Successively we can derive all bits of product $C$ using the same function $f_{mult}$, and the only
adjustment we need to make is to right-cyclic shift 2 operands by 1 bit each time.
\end{Example}

@@

If we want to make the complexity of $f_{mult}$ lower than $O(k^2)$, then the best choice is to try out 
linear functions. As we know, matrix multiplication can simulate all possible combinations of 
linear functions. 
% (which is also the reason it is used as basic model to simulate the behavior of a neuron
% in neural network machine learning algorithms).
Imagine $A$ is a $k$-bit row vector and $B$ is a $k$-bit 
column vector, then the single bit product can be written as the product of matrix multiplication
$$c_{l} = A\times M\times B$$
where 
\begin{align*}
A &= (a_0,a_1,\dots,a_{k-1})\\
B &= \begin{pmatrix}
b_0 \\
b_1 \\
\vdots \\
b_{k-1}
\end{pmatrix}
\end{align*}
Moreover, $M$ is a $k\times k$ square binary matrix. If we can find $M$, we obtain the design of the multiplier.

@@

When taking different bits $l$ of the product in Equation \ref{eqn:def_lambda}, 
we obtain a series of conjugate matrices of $M$. 
Which means instead of shifting operands $A$ and $B$, we can also shift the 
matrix.

More specifically, we denote the matrix by \emph{$l$-th $\lambda$-Matrix} as 
$$c_l = A \times M^{(l)} \cdot B^T$$
Meanwhile, the operator shifting rule in Equation \ref{eqn:shiftMSB} still holds. Then we have the relation 
$$c_{l-1} = A \cdot M^{(l-1)} \cdot B^T = shift(A) \cdot M^{(l)} \cdot shift(B)^T$$ which means
by right and down cyclically shifting $M^{(l-1)}$, we can get $M^{(l)}$.

@@

If we generalize the definition and explore the nature of $\lambda$-Matrix, it is defined as cross-product terms from multiplication, which is 
\begin{equation}
Product~vector~C = (\sum_{i=0}^{k-1}a_i\beta^{2^i})(\sum_{j=0}^{k-1}b_j\beta^{2^j}) = \sum_{i=0}^{k-1}\sum_{j=0}^{k-1}a_ib_j\beta^{2^i}\beta^{2^j}
\end{equation}
The expressions $\beta^{2^i}\beta^{2^j}$ are referred to as cross-product terms, and can be represented by
NB, {\it i.e.}
\begin{equation}
\beta^{2^i}\beta^{2^j} = \sum_{l=0}^{k-1}\lambda_{ij}^{(l)}\beta^{2^l}, \ \ \lambda_{ij}^{(l)} \in \mathbb F_2.
\end{equation}
Substitution yields an expression for $l$-th digit of product as showed in Equation \ref{eqn:multiMSB}:
\begin{equation}
c_l = \sum_{i=0}^{k-1}\sum_{j=0}^{k-1}\lambda_{ij}^{(l)}a_ib_j
\end{equation}
$\lambda_{ij}^{(l)}$ is the entry with coordinate $(i,j)$ in $l$-th $\lambda$-Matrix.

The $\lambda$-Matrix can be implemented with XOR and AND gates in circuit design.
The very naive implementation requires $O(C_N)$ gates, where $C_N$ is the number of
nonzero entries in $\lambda$-Matrix.
There usually exists multiple NBs in $\Fkk, k>3$. If we employ a random NB, there is no mathematical
guarantee that $C_N$ has bound $o(k)$. However,
Mullin {\it et al.} \cite{mullinONB} % This citation is valid
proves that in certain GF $\mathbb F_{p^{k_{opt}}}$, there always exists at least one NB such that 
its corresponding $\lambda$-Matrix has $C_N = 2n-1$ nonzero entries. A basis with this property is
called optimal normal basis (ONB), details are introduced in Appendix \ref{append:ONB}.

@@

% After fixing this, add a whole-piece StdB vs NB example, list their cost as the conclusion
\subsection{Comparison between Standard Basis and Normal Basis}
At the end of this section, a detailed example is used to make a comparison between StdB multiplication
and NB multiplication.
\begin{Example}[Rijndael's finite field]
\label{ex:Rijndael}
Rijndael uses a characteristic 2 finite field with 256 elements, which can also be called the GF $\mathbb F_{2^8}$.
Let us define the primitive element $\alpha$ using irreducible polynomial 
$\alpha^8+\alpha^7+\alpha^6+\alpha^4+\alpha^2+\alpha+1$. Coincidently, $\alpha$ is also a normal element,
i.e. $\beta = \alpha$ can construct a NB $\{\alpha,\alpha^2,\alpha^4,\alpha^8,\alpha^{16},\alpha^{32},\alpha^{64},
\alpha^{128}\}$.

@@

The final remainder is $1$, {\it i.e.} the product equals to 1 in StdB.
